Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-21 latex/intro.tex
I think its donegit add why.tex

\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates $\obs$
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  code in supplementary material). Given our model's genericity, we
  expect that measurement clusterization to be a generic phenomenon.

@@

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:why}, we present a compelling
  explanation for the optimality of clustered designs in the absence
  of model error. Our analysis reveals that, in our model, a D-optimal
  design focuses on a select set of prior eigenvectors, specifically
  those with the highest power spectrum. In practical scenarios, the
  number of locations where the relevant prior eigenvectors are
  significantly large, while other eigenvectors have a power spectrum
  of zero, is limited. Consequently, the clustering of measurements
  arises as a natural consequence of the pigeonhole principle, as
  there are more measurements available than there are locations
  satisfying conditions (a) and (b).

@@

Lastly, in Lemma \ref{lemma:free} we decompose a symmetric positive
definite matrix $M \in \mathbb{R}^{k \times k}$ with $\ttr M = m \in
\mathbb{N}$ as $M = AA^t$, where $A$ has unit norm columns.

>>>>

\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates $\obs$
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  code in supplementary material). Given our model's genericity, we
  expect measurement clusterization to be a generic phenomenon.

@@

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:why}, we present a compelling
  explanation for the optimality of clustered designs in the absence
  of model error. Our analysis reveals that, in our model, a D-optimal
  design focuses on a select set of prior eigenvectors, specifically
  those with the largest eigenvalues in the prior covariance
  spectrum. In practical scenarios, the number of locations where the
  relevant prior eigenvectors are significantly large, while other
  eigenvectors have a power spectrum close to zero, is
  limited. Consequently, the clustering of measurements arises as a
  natural consequence of the pigeonhole principle, as there are more
  measurements available than there are locations satisfying
  conditions (a) and (b).

@@

In the proving Theorem \ref{thm:char} we prove and generalize several
lemmas. Among those, is a Lemma \ref{lemma:free}, which is (to the the
best of my knowledge) a novel lemma in linear algebra: We decompose a
symmetric positive definite matrix $M \in \mathbb{R}^{k \times k}$
with $\ttr M = m \in \mathbb{N}$ as $M = AA^t$, where $A$ has unit
norm columns.

### 2023-05-21 latex/paper.tex
I think its donegit add why.tex

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  Inversion for parameteres of physical processes typically requires
  taking expensive measurements; finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal designs are sometimes extremely
  clustered. Since measurement clusterization is nonintuitive,
  researchers often avoid it by modifying the optimal design
  problem. We consider a certain flavor of the optimal design problem,
  based on the Bayesian D-optimality criterion, and suggest an
  analytically tractable model for Bayesian D-optimal designs in
  linear inverse problems over Hilbert spaces. We prove that
  measurement clusterization poses no obstruction to D-optimality. We
  demonstrate that measurement clusterization is a generic property of
  D-optimal designs, and prove that correlated noise between
  measurements mitigates clusterization. We also give a full
  characterization of D-optimal designs, which shows that D-optimal
  designs uniformly reduce uncertainty in the space of
  observations. Finally, we show how measurement clusterization can be
  avoided without changing the optimal design problem.
\end{abstract}

@@

\section{Acknowledgements}
This study is a result of research I started during my PhD studies
under the instruction of Prof. Georg Stadler at the Courant Institute
of Mathematical Sciences. I would like to thank him for his great
mentorship, attention to details and kindness. This work was also
supported by The Raymond and Beverly Sackler Post-Doctoral
Scholarship.

This research was supported in part by an appointment with the
National Science Foundation (NSF) Mathematical Sciences Graduate
Internship (MSGI) Program sponsored by the NSF Division of
Mathematical Sciences. This program is administered by the Oak Ridge
Institute for Science and Education (ORISE) through an interagency
agreement between the U.S. Department of Energy (DOE) and NSF. ORISE
is managed for DOE by ORAU. All opinions expressed in this paper are
the author's and do not necessarily reflect the policies and views of
NSF, ORAU/ORISE, or DOE.

>>>>

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  Inversion for parameters of physical processes typically requires
  taking expensive measurements; finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal designs are sometimes extremely
  clustered. Since measurement clusterization is nonintuitive,
  researchers often avoid it by modifying the optimal design
  problem. We consider a certain flavor of the optimal design problem,
  based on the Bayesian D-optimality criterion, and suggest an
  analytically tractable model for Bayesian D-optimal designs in
  linear inverse problems over Hilbert spaces. We demonstrate that
  measurement clusterization is a generic property of D-optimal
  designs, and prove that correlated noise between measurements
  mitigates clusterization. We also give a full characterization of
  D-optimal designs, which shows that D-optimal designs uniformly
  reduce uncertainty in a select subset of the prior covariance
  eigenvectors. Finally, we show how measurement clusterization is a
  simple consequence of the characterization mentioned above and the
  pigeonhole principle.
\end{abstract}

@@

\section{Acknowledgements}
This study is a result of research I started during my PhD studies
under the instruction of Prof. Georg Stadler at the Courant Institute
of Mathematical Sciences. I would like to thank him for his great
mentorship, attention to details and kindness. I would also like to
thank Christian Remling, who helped me find a proof for Lemma
\ref{lemma:free} in
\href{https://mathoverflow.net/questions/280168/redistribute-diagonal-entries-of-a-matrix/280203#280203c}{Mathoverflow}.

This research was supported in part by an appointment with the
National Science Foundation (NSF) Mathematical Sciences Graduate
Internship (MSGI) Program sponsored by the NSF Division of
Mathematical Sciences. This program is administered by the Oak Ridge
Institute for Science and Education (ORISE) through an interagency
agreement between the U.S. Department of Energy (DOE) and NSF. ORISE
is managed for DOE by ORAU. All opinions expressed in this paper are
the author's and do not necessarily reflect the policies and views of
NSF, ORAU/ORISE, or DOE. This work was also supported by The Raymond
and Beverly Sackler Post-Doctoral Scholarship.

### 2023-05-21 latex/why.tex
I think its donegit add why.tex

\section{An answer for Question \ref{q:why}: The cause of measurement clusterization}\label{section:why}
\begin{figure}[H]
    \centering
    \includegraphics[height=0.5\textwidth]{eigenvectors.eps}
    \caption{D-optimal measurement locations ($m=5$ measurements) and
      weighted eigenvectors for inversion of the initial condition of
      the 1D heat equation (see supplementary material for
      details). Measurement locations and weighted eigenvectors are
      plotted over the computational domain $\Omega = [0, 1]$
      (x-axis).  Measurement clusterization occurs approxmately at
      $0.31$ and $0.69$. These two locations are a compromise between
      zerosz of the irrelevant eigenvectors (third and up) and staying
      far from a zero of the first and second eigenvectors.}
  \label{fig:why}
\end{figure}

According to Theorem \ref{thm:char}, D-optimal designs aim to measure
a small subset of eigenvectors; namely, eigenvectors $1$ through
$k$. In our model, not measuring eigenvectors $k+1$ and above is
readily achieved, as the proof of Theorem \ref{thm:char} demonstrates.
Taking this understanding to spatial problems, we expect a D-optimal
design to prefer measurement locations where eigenvectors $k+1$ and
above are (close to) zero: either because their eigenvalue in the
spectrum of the prior is small, or because the value of the
eigenvector itself is (close to) zero. Figure \ref{fig:why}
demonstrates this preference for the 1D heat equation with homogeneous
Dirichlet boundary conditions (details in the supplementary
material). We plot four eigenvectors, scaled according to their prior
standard deviations. We ignore eigenvectors above the fourth, due to
their small prior eigenvalues. We can observe that measurements are
clustered close to zeros of the third and fourth eigenvectors. The
reason for this measurement clusterization is that there are only two
locations where the third and fourth eigenvectors are close to zero,
and the first and second eigenvectors are considerably larger than
zero. When we fit four measurements into these measurement locations,
we find that measurements cluster, in accordance with the pigeonhole
principle.

%% \textbf{Question \ref{q:why}}:

>>>>

\section{An answer for Question \ref{q:why}: The cause of measurement clusterization}\label{section:why}
\begin{figure}[H]
    \centering
    \includegraphics[height=0.5\textwidth]{eigenvectors.eps}
    \caption{D-optimal measurement locations ($m=5$ measurements) and
      weighted eigenvectors for inversion of the initial condition of
      the 1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between zeros of eigenvectors a D-optimal design aims
      to ignore (third and up) and staying far from a zero of the
      first and second eigenvectors. Allocating $m=5$ measurements
      into two locations results in clusterization, according to the
      pigeonhole principle.}
  \label{fig:why}
\end{figure}

According to Theorem \ref{thm:char}, D-optimal designs aim to capture
a small subset of eigenvectors of the prior covariance, specifically
the $k$ eigenvectors with the highest prior variance. Our model
naturally achieves this objective by not measuring eigenvectors $k+1$
and above, as proven in Theorem \ref{thm:char}. Translating this
understanding to spatial problems, we anticipate that a D-optimal
design would favor measurement locations where eigenvectors $k+1$ and
above are either close to zero in value or possess small eigenvalues
in the prior spectrum, for some $k > 0$. To illustrate this
preference, Figure \ref{fig:why} depicts the scenario using the 1D
heat equation with homogeneous Dirichlet boundary
conditions\footnote{Additional details provided in the supplementary
material.}. The plot showcases four eigenvectors, scaled according to
their prior standard deviations. Since eigenvectors beyond the fourth
have insignificant prior eigenvalues, we exclude them from
consideration. Notably, we observe that measurements are clustered
near the zeros of the third and fourth eigenvectors, so we conclude
that $k=2$. Clusterization arises because there are only two locations
where the third and fourth eigenvectors approach zero, while the first
and second eigenvectors exhibit significantly non-zero
values. Consequently, when we allocate four measurements to these two
locations, they naturally cluster together, aligning with the
pigeonhole principle.
